---
title: "Wages and Labor Force Participation in the US"
author: "Yukako Kitamura"
output: pdf_document
fontsize: 11pt
urlcolor: blue
---

The main goal of this study is to find out how hourly wages and labor force participation (LFP) have evolved for skilled and unskilled workers in the United States. The dataset used is the U.S. Current Population Survey between 1976 and 2015, which is extracted from the IPUMS. Following the explanation on their website found [here](https://cps.ipums.org/cps/cpi99.shtml), I obtained CPI99 values and converted the nominal wages in the data to 2017 dollars to allow the historical comparison in wages.

```{r setup, include=F}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(gridExtra)
```

```{r dataloading, message = F, warning = F, cache = T}
cps <- read_csv("cps_wages_LFP.csv")
```

```{r preparation, cache = TRUE}
#Using definition of the labor force, I exclude people who are under age 16
#Also, I create dummy for unemployment and labor force for the easier data handling
#Note: there are a lot of observations with 0 wage regardless of their positive inctot.
#      However, excluding obs with 0 hourly wage and leaving obs with 0.0002 hourly wage
#      doesn't make much sense, so I will use all values to calculate the mean unless they 
#      have NA here

cps_lf <- cps %>% mutate(age = as.numeric(age)) %>%
                  filter(age > 15) %>%
                  mutate(unemployed = case_when(empstatid == "Employed" ~ 0,
                                                empstatid == "Unemployed" ~ 1,
                                                TRUE ~ as.numeric(NA)),
                         lfp = ifelse(lfp == "In labor force", 1, 0),
                         sex = as.logical(sex == "male")) %>%
                  rename(weight = wtsupp)

#Based on the deifinition of the labor force, I include armed forces into LFP 
#(originally having missing values)
cps_lf$lfp[is.na(cps_lf$lfp)] <- 1

#Create education category for college experiences - no(high school), some, or
#BA and over. For the education measurement before 1992, I consider 4 years of college 
#equivalent to BA
educlist <- unique(cps$educ)
cps_lf <- cps_lf %>% mutate(educ2 = case_when(
                                    is.na(educ) ~ "NA",
                                    educ %in% educlist[c(1,3:6,9:15,17,20,22,25:27)] ~ "no",
                                    educ %in% educlist[c(8,16,21,23,30,32)] ~ "some",
                                    TRUE ~ "BA"))
rm(educlist)
```

```{r yearlydata}
#Aggregate data by year
yearly <- cps_lf %>%
          mutate(BA = ifelse(educ2 == "BA", 1, 0)) %>%
          group_by(year) %>%
          summarize(LFP_all = weighted.mean(lfp, w = weight),
                    LFP_male = weighted.mean(lfp[sex == 1], w = weight[sex == 1]),
                    LFP_female = weighted.mean(lfp[sex == 0], w = weight[sex == 0]),
                    LFP_under25 = weighted.mean(lfp[age_group == "age < 25"], 
                                                w = weight[age_group == "age < 25"]),
                    LFP_under45 = weighted.mean(lfp[age_group == "25 <= age < 45"], 
                                                w = weight[age_group == "25 <= age < 45"]),
                    LFP_under65 = weighted.mean(lfp[age_group == "45 <= age < 65"], 
                                                w = weight[age_group == "45 <= age < 65"]),
                    LFP_over65 = weighted.mean(lfp[age_group == "65 <= age"], 
                                               w = weight[age_group == "65 <= age"]),
                    unemp = weighted.mean(unemployed, w = weight, na.rm = T),
                    f_BA = weighted.mean(BA, w = weight),
                    wage_m = weighted.mean(wage[sex == 1], w = weight[sex == 1], na.rm = T),
                    wage_f = weighted.mean(wage[sex == 0], w = weight[sex == 0], na.rm = T),
                    wage_all = weighted.mean(wage, w = weight, na.rm = T))

#Get CPI99 values from here (https://cps.ipums.org/cps/cpi99.shtml) to take inflation into 
#account. Using these values, I convert nominal hourly wage to 1999 dollars first (nominal 
#wage*CPI99) and then to 2017 dollars by multiplying 1.471
cpi99 <- c(2.928, 2.749, 2.555, 2.295, 2.022, 1.833, 1.726, 1.673, 1.603, 1.548,
           1.520, 1.467, 1.408, 1.344, 1.275, 1.223, 1.187, 1.153, 1.124, 1.093, 1.062,
           1.038, 1.022, 1.000, 0.967, 0.941, 0.926, 0.905, 0.882, 0.853, 0.826, 0.804,
           0.774, 0.777, 0.764, 0.741, 0.726, 0.715, 0.704, 0.703)
yearly <- cbind(yearly, cpi99) %>%
          mutate(wage17_m = wage_m*cpi99*1.471, wage17_f = wage_f*cpi99*1.471,
                 wage17_all = wage_all*cpi99*1.471)
```

# A. Overall Trend between 1976 and 2015

To understand the overview of historical transition, I first summarize the LFP and hourly wages by gender as well as the other relevant information in Table1. The first notice is the decline in the male LFP. Despite the steady increase in the wages, male LFP has been decreasing, whereas the female has experienced increases in both the LFP and wages. It is interesting to note that even in the recession, which is shown by the higher unemployment rate in 2010, the wage seems to remain stable on average. Also, a huge increase in education level over years is clearly expressed by the ratio of people who hold at least bachelor degrees in the last column.

&nbsp;
```{r table1}
#This is for the table summarizing the overall trend
a <- yearly %>% filter(year %in% c(seq(1980, 2015, by = 5))) %>%
                select(year, LFP_male, LFP_female, wage17_m, wage17_f, unemp, f_BA)


colnames(a) <- c("Year", "Male", "Female", "Male", "Female", "Unemployed", "Bachelor")
kable(a, "latex", booktabs = T, linesep = "", caption = "Trend Summary", align = "c",
      digits = 3) %>%
        add_header_above(c(" ", "LFP" = 2, "Hourly Wage" = 2, "Ratio" = 2)) %>%
        kable_styling(position = "center", font_size = 11, 
                      latex_options = "hold_position", full_width = F) %>%
        column_spec(1, width = "1cm") %>%
        column_spec(2:7, width = "1.8cm") %>%
        add_footnote(c("Hourly wages are expressed in 2017 dollars.",
                       "LFP = (Civilian labor force+Armed forces)/Total population 16 and over",
                       "The unemployed ratio = Unemployed/Civilian labor force"),
                     notation = "number")
rm(a)
```
\pagebreak

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; The next plot, Chart1, shows the change in LFP by the different age groups. The obvious decline in the LFP for people under 25 can be explained by the fact I have seen in Table1; that is, people are more likely to stay in a school longer in the recent years, and thus, the ratio of young people staying in labor force has largely decreased. The LFP of prime working-age people (25 - 64) slightly increased between 1976 and 1995, but it has got almost steady in the recent 20 years. On the contrary, LFP of people aged over 65 has been increasing in the recent 20 years. This implies that the retirement age is getting older, reflecting the aging society.  

&nbsp;  
&nbsp;  
```{r chart1}
#For the chart describing the change in LFP by age group
LFP <- yearly %>% select(year, "16 to 24" = LFP_under25, "25 to 44" = LFP_under45, 
                         "45 to 64" = LFP_under65, "Over 65" = LFP_over65) %>%
                  gather(key = name, value = value, '16 to 24', '25 to 44',
                         '45 to 64', 'Over 65')


ggplot(LFP, aes(x = year, y = value, color = name)) + geom_line() +
    labs(color = "Age Group") + ggtitle("Chart1: Labor Force Participation by Age") +
    ylab("Rate of LFP")  + xlab("Year") + theme_bw() + 
    theme(plot.margin= unit(c(1, 0, 1, 0), "cm"), 
          plot.title = element_text(face = "italic", size = 11, hjust = 0.5))
rm(LFP)
```

```{r preparation2}
#Here, I will only use male aged between 25 and 64 (prime working-age people) and obtain
#the rate for each skilled group (skill is defined by the education level)
over25 <- cps_lf %>%
          filter(age >= 25 & age < 65 & sex == 1) %>%
          group_by(year) %>%
          summarize(w_skilled = weighted.mean(wage[educ2 == "BA"],
                                        w = weight[educ2 == "BA"], na.rm = T),
                    w_semiskilled = weighted.mean(wage[educ2 == "some"],
                                        w = weight[educ2 == "some"], na.rm = T),
                    w_unskilled = weighted.mean(wage[educ2 == "no"],
                                                w = weight[educ2 == "no"], na.rm = T),
                    LFP_skilled = weighted.mean(lfp[educ2 == "BA"], 
                                          w = weight[educ2 == "BA"]),
                    LFP_semiskilled = weighted.mean(lfp[educ2 == "some"], 
                                          w = weight[educ2 == "some"]),
                    LFP_unskilled = weighted.mean(lfp[educ2 == "no"], 
                                         w = weight[educ2 == "no"]))

#Convert wages to 2017 dollars
over25 <- cbind(over25, cpi99) %>%
          mutate(w_skilled = w_skilled*cpi99*1.471, 
                 w_semiskilled = w_semiskilled*cpi99*1.471,
                 w_unskilled = w_unskilled*cpi99*1.471)
```
&nbsp;

# B. Skills, Wages, and Labor Force Participation

In this section, I will focus on one's skill level defined by his/her education level and examine its influence on a wage and LFP. Expanding the dataset's original classification which is to see if one has more than a high school diploma, including one year of college experience, I will define skill levels in the following way:  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Skilled - bachelor degree or more;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Semiskilled - more than a high school diploma, but not completing a college;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Unskilled - high school degree or less  
The reason for adding the subcategory is there is a huge difference between the completion of bachelor and one year of college experience, especially in the context of a huge increase in BA ratio. In other words, in the society where one-fourth of people have bachelor degrees, the skills gained from a year of college experience are valued much less than a couple of decades ago when going and completing college was rarer.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Also, since male and female have different trends, I will only use male and focus on those who are in their prime working-age (25 - 64) not to be distracted by any change driven by the society's systematic change: namely, more schooling and aging. Another reason for excluding old people is they might have some age-specific differences between skilled and unskilled people, led by a factor correlating with education. An example is such factor is health. Since educated people are more likely to be healthy, they could work longer; or more simply, there might be much more people who are over 80, which systematically decreases the LFP. Therefore, I decided to compare people who are in relatively the same condition across different skill levels.  
&nbsp;  

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; The next two plots show the change in wage and LFP by different skill groups. In the Chart2, we can see the tremendous increase in the hourly wage of skilled people as opposed to the flat wages for lower-skilled people. This implies that an increase in overall wages over years we saw in Table1 is mainly caused by an increase in a wage of the skilled people and that there is a huge disparity between bachelor holders and the others.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; On the other hand, Chart3 shows that the changes in LFP are much more significant for lower skilled people than skilled people. Compared to roughly 5% of a decrease in LFP rate for skilled people, both semiskilled and unskilled people have experienced roughly 10% of a decrease. Combining these two results, it can be said that much lower relative wage for lower-skilled people discourages them and pushes them out of labor force.  

&nbsp;  

```{r chart2_chart3}
#Making charts for the change in wages and LFP by different skill group
wage <- over25 %>% select(year, Skilled = w_skilled, Semiskilled = w_semiskilled, 
                          Unskilled = w_unskilled) %>%
                   gather(key = group, value = value, Skilled, Semiskilled, Unskilled) %>%
                   mutate(group = 
                           factor(group, levels = c("Skilled", "Semiskilled", "Unskilled")))
  
LFP <- over25 %>% select(year, Skilled = LFP_skilled, Semiskilled = LFP_semiskilled,
                         Unskilled = LFP_unskilled) %>%
                  gather(key = group, value = value, Skilled, Semiskilled, Unskilled) %>%
                  mutate(group = 
                          factor(group, levels = c("Skilled", "Semiskilled", "Unskilled")))

ggplot(wage, aes(x = year, y = value, color = group)) + geom_line() +
    scale_color_discrete(name = "Skill group") +
    ggtitle("Chart2: Change in Wage for Male Aged 25 - 64") + 
    ylab("Wage in 2017 Dollars") + xlab("Year") + 
    theme_bw() + theme(plot.margin= unit(c(1, 0, 1, 0), "cm"), 
          plot.title = element_text(face = "italic", hjust = 0.5, size = 11))


ggplot(LFP, aes(x = year, y = value, color = group)) + geom_line() +
     scale_color_discrete(name = "Skill group") +
     ggtitle("Chart3: Change in Labor Force Participation Rate for Male Aged 25 - 64") + 
     ylab("Rate of LFP")  + xlab("Year") + 
     theme_bw() + theme(plot.margin= unit(c(1, 0, 1, 0), "cm"), 
          plot.title = element_text(face = "italic", hjust = 0.5, size = 11))
rm(wage, LFP)
```
&nbsp;  
&nbsp;

# C. Discussion

## Theoretical Hypothesis

In the last section, a possible explanation about the different pattern of LFP transition between skilled and low-skilled people is revealed through the difference in the wage transition between two groups. However, it is not still clear what has caused a significant rise in skilled people's wages. To explore this question, I will consider a possible mechanism of the change in two labor markets for skilled and low-skilled people through the model of labor supply and demand.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Below, I drew two hypothetical graphs for the skilled and unskilled labors. The main assumption made here is the elasticity of labor demand. To be specific, I assume demand for unskilled labor is much more elastic than the one for skilled labor because low-skilled jobs can be much more easily substituted by other workers than skilled jobs. I also assume these two graphs to show *relative* supply and demand. Put differently, I will consider the relative supply and demand for skilled labor to unskilled labor in the first graph, and vice versa in the second graph.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Based on what I have found from the dataset, I will first think about the market for skilled labor. The solid two lines show the original demand and supply. From here, a *relative* increase in skilled labor to lower-skilled labor, which is correspondent to the increase in bachelor holders' ratio, pushes the supply curve right to the black dotted line. If this is the only phenomenon that had happened, the wage must have been lower than the original wage which is the intersection of the new supply curve and the original demand curve. However, what has actually happened is an increase in the relative wage of skilled labor. Thus, to explain this situation with this graph, we need to make a hypothesis that the relative demand for skilled jobs has also significantly increased as expressed in the red dotted line, leading a higher relative wage for the skilled labor. Likewise, if I consider there has been a decrease in *relative* demand for unskilled labor, the lower *relative* wage for them can be interpreted as the consequence of a decrease in *relative* supply of unskilled labor.  
To sum up, this theoretical overview has shown me that a different pattern of wage transition for skilled and unskilled labor can be explained by an increase in relative demand for skilled jobs along with an increase in skilled labor.  
&nbsp;  


```{r}
#The following code is to produce the theoretical demand and supply curve. I adjusted the 
#number just to support my explanation, so the numbers are completely arbitrary and does not #have any mean
par(mfrow = c(1, 2), oma = c(3,0,3,0), mai = c(0.8,0.3,0.4,0.3))
plot(0, type = "n", xlab = "", ylab = "", xaxt = 'n', yaxt = 'n',
     bty = 'l', main = "Skilled Labor")
segments(0.7,-0.75,1.2,0.8)
segments(0.85,-0.75,1.35,0.8, lty = "dashed", col = "black")
segments(1,-0.75,0.9,0.8)
segments(1.25,-0.75,1.15,0.8, lty = "dashed", col = "red")
points(0.95, 0.025, pch = 16)
points(1.185,0.28, pch = 16, col = "red")
text(0.83,0.025, "original wage", cex = 0.6, col = "black")
text(1.28,0.28, "new wage", cex = 0.6, col = "red")
text(1.03,-0.65, "D", cex = 0.8, col = "black")
text(1.22,0.7, "S", cex = 0.8, col = "black")
text(0.53, 0.255, cex = 0.75,"Wage", srt = 90, xpd = T)
par(xpd = T)
legend(1, -1.11, cex = 0.75, legend = "Quantity of Labor", bty = "n")

plot(0, type = "n", xlab = "", ylab = "", xaxt = 'n', yaxt = 'n',
     bty = 'l', main = "Unskilled Labor")
segments(0.7,-0.75,1.2,0.8, lty = "dashed", col = "black")
segments(0.85,-0.75,1.35,0.8)
segments(1.4,-0.75,0.7,0.8)
segments(1.1,-0.75,0.6,0.35, lty = "dashed", col = "red")
points(1.08,-0.04, pch = 16)
points(0.866,-0.228, pch = 16, col = "red")
text(1.2,-0.04, "original wage", cex = 0.6, col = "black")
text(0.755,-0.228, "new wage", cex = 0.6, col = "red")
text(1.28,-0.65, "D", cex = 0.8, col = "black")
text(1.33,0.53, "S", cex = 0.8, col = "black")
text(0.53, 0.255, cex = 0.75,"Wage", srt = 90, xpd = T)
par(xpd = T)
legend(1.015, -1.5, cex = 0.6,
       legend = c("New Labor Demand", "New Labor Supply"), col = c("black", "red"),
       lty = c("dashed", "dashed"))
legend(1, -1.11, cex = 0.75, legend = "Quantity of Labor", bty = "n")
```
&nbsp;  

## Further Investigation

To further investigate this question and find more evidence to support the hypothesis I have made above, there are three steps I can think of. First, examining how the industry structure has changed over years and if the wage disparity has expanded across sectors will give us an overview of the change in labor demand. Second, comparing LFP and minimum wage at the state level will be beneficial to support the idea that a low wage is the main determinant to push lower skilled people out of the labor force. The third step is to obtain data at the regional level such as PUMA because sufficient variation in educational attainment across PUMAs will allow us to closely analyze the relationship between education and other labor market outcomes. For example, we may see an evidence that a region that has experienced a huge increase in the educational attainment might have experienced a large expansion of technical jobs, which leads to a higher demand for skilled labor. Moreover, looking at unskilled people's LFP with respect to the region's educational attainment could be an interesting point at the regional level as well.